---
title: "Data Project Part 3: Exploring relationship between exposure to Uranium and Renal Disease"
author: "Castle Funatake, Vicki Ip, Meredith Potts, Matthew Reyes, Anastasia Roloff"
date: "04/27/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1 Questions

1. What is the problem your are addressing with these data? State the question you are trying to answer and let us know what type of question this is in terms of the PPDAC framework.

The problem that is being addressed with this dataset is whether chronic ingestion of uranium in drinking water affects kidney function. In terms of the PPDAC framework, this would be a cause and effect type, or etiologic, question. Uranium concentration in water is the ordinal exposure variable and has been quantified as either low exposure (<1 microgram/L) or high exposure (2-781 microgram/L), with the chronicity of exposure measured in terms of years of residence – between 1 and 33 years in the low exposure group, and 3 to 59 years in the high exposure group. Uranium concentration in urine was also measured as a continuous exposure variable. The response variables correspond to kidney function indicators (glucose, creatinine, protein, beta-2-microglobulin (BMG)) and markers of cell toxicity (ALP, GGT, LDH, NAG), which were all sampled via urine.

\newpage

2. What is the target population for your project? Why was this target chosen i.e., what was your rationale for wanting to answer this question in this specific population?

The target population for this study is adults at risk for exposure to uranium through drinking water. The study population was selected from a village in Nova Scotia known to have elevated levels of uranium in water from private wells and a group of healthy volunteers from Ottawa, Ontario who received municipally supplied drinking water.  All subjects, regardless of location, were grouped according to levels of uranium in their drinking water. 
Our group is concurrently working on a project in 200F studying the effects of uranium mining on the Navajo Nation. The land of the Navajo is naturally rich in uranium, thus, despite having been a major source of uranium mining, many residents living far from abandoned mines are still exposed to naturally-occurring uranium in their drinking water as a majority are reliant on unregulated water sources. Although the results from this dataset cannot be extrapolated onto Navajo residents, it does provide some insight into whether differing levels of uranium exposure affect kidney function in humans.


\newpage

3. What is the sampling frame used to collect the data you are using? Describe why you think this sampling strategy is appropriate for your question. To what group(s) would you feel comfortable generalizing the findings of your study and why.

The sampling population were adult volunteers aged 13-87 either from a village in Nova Scotia with private wells that had uranium content above the Canadian guidelines of 100 microgram/L or from Ottawa, Ontario with access to treated municipal water with uranium levels less than 1 microgram/L. The baseline uranium levels were collected at the beginning of the study through sampling of tap and well water. The authors attempted to have an even distribution of sex and ages by matching healthy participants from Ottawa to those from Nova Scotia. 24 hour urinary samples were taken from each participant and measured for levels of Uranium indicators of kidney function and nephrotoxicity. The data from participants was grouped according to low and high uranium exposure, regardless of place of residence or source of drinking water. Both urine levels of uranium were analyzed along with concentrations of uranium in local drinking water supplies. The most recent period of exposure was used as the exposure window for drinking water concentrations, since wells were remediated over time. The outcomes measured were biomarkers of kidney function such as urine volume, microglobulin, glucose, and albumin, as well as specific enzymes to determine local cell damage.

Ideally this sample population could be representative of any target population of adults exposed to high levels of uranium through drinking water, however the study sample is small (50 participants) and no data were collected on socioeconomic status or other demographic characteristics, making it difficult to generalize the findings beyond the adult populations in these two cities in Canada. Consequent uranium levels were measured through duplicate portions of food and water that each participant consumed.  
This sampling strategy is appropriate for the question of the study because they were trying to measure kidney effects due to chronic uranium exposure in differing levels of drinking water, which in theory should be similar across humans based on the toxicokinetics of uranium.  However, including more participant characteristics and disaggregated data could help the study be more generalizable to a wider population.


\newpage

4. Write a brief description (1-4 sentences) of the source and contents of your dataset. Provide a
URL to the original data source if applicable. If not (e.g., the data came from your internship), provide
1-2 sentences saying where the data came from. If you completed a web form to access the data and
selected a subset, describe these steps (including any options you selected) and the date you accessed
the data

The dataset was obtained from the study by Zamora et al. (1998):


Zamora, M. L., Tracy, B. L., Zielinski, J. M., Meyerhof, D. P., & Moss, M. A. (1998). Chronic ingestion of uranium in drinking water: A study of kidney bioeffects in humans. Toxicological Sciences: An Official Journal of the Society of Toxicology, 43(1), 68–77. https://doi.org/10.1006/toxs.1998.2426.



\newpage

5. Write code below to import your data into R. Assign your dataset to an object.

```{r importing-data, warning=FALSE, echo=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
nn_data <- read.csv("NNdata.csv")
head(nn_data)
```
\newpage

6. Use code in R to answer the following questions:

- What are the dimensions of the dataset?
```{r dimensions, warning=FALSE, echo=FALSE}
dim(nn_data)
```

- Provide a list of variable names.
```{r variable-names, warning=FALSE, echo=FALSE}
vars <- names(nn_data)
vars
```

- Print the first six rows of the dataset.

```{r head, warning=FALSE, echo=FALSE}
firstSix <- head(nn_data)
firstSix
```

\newpage

7. Use the data to demonstrate a statistical concept from Part I of the course. Describe the concept that you are demonstrating and interpret the findings. This should be a combination of code and written explanation.

First to explore the data, we checked the mean values for each of the variables, grouped by exposure level (`Exp_level`). Below is the summary table with those findings:

```{r summary-tables, warning=FALSE, echo=FALSE}
# factor Exposure level
nn_data$Exp_Level <- as.factor(nn_data$Exp_Level)
summary <- nn_data %>%
  group_by(Exp_Level) %>%
  summarize(
            uranium_mean = mean(U_intake),
            uranium_median = median(U_intake),
            uranium_sd = sd(U_intake),
            glucose_mean = mean(Glucose),
            glucose_median = median(Glucose),
            glucose_sd = sd(Glucose),
            creatinine_mean = mean(Creatinine),
            creatinine_median = median(Creatinine),
            creatinine_sd = sd(Creatinine),
            protein_mean = mean(Protein),
            ALP_mean = mean(ALP),
            GGT_mean = mean(GGT),
            LDH_mean = mean(LDH),
            NAG_mean = mean(NAG),
            )
summary
summary_ungrouped <- nn_data %>%
  summarize(
            uranium_mean = mean(U_intake),
            uranium_median = median(U_intake),
            uranium_sd = sd(U_intake),
            glucose_mean = mean(Glucose),
            glucose_median = median(Glucose),
            glucose_sd = sd(Glucose),
            creatinine_mean = mean(Creatinine),
            creatinine_median = median(Creatinine),
            creatinine_sd = sd(Creatinine),
            protein_mean = mean(Protein),
            ALP_mean = mean(ALP),
            GGT_mean = mean(GGT),
            LDH_mean = mean(LDH),
            NAG_mean = mean(NAG),
            )
summary_ungrouped
# G_mean = 96.08, sd = 80.4045, C_mean = 9.712, sd = 3.415
```

Next, we visualized the data by graphing using a scatterplot the observed urine uranium concentration levels (`U_intake`) with glucose (`Glucose`) as a response variable. We also grouped the individual data points to visualize if there was any discrepancy in the relationship based on Sex (`Sex`). That scatter plot is depicted below:

```{r scatterplot1, warning = FALSE, echo = FALSE}
sc_plot <- ggplot(nn_data, aes(x= U_intake, y = Glucose)) +
  geom_point(aes(col = Sex)) +
  theme_minimal(base_size = 15)
sc_plot
```

Next, we plotted a histogram of the same uranium intake concentration. Noticing that the below histogram was right-skewed and had a tall unimodal peak at the lowest concentrations of uranium, we then did a log transformation of that variable and subsequently visualized the initial histogram of uranium concentration, except now using the log-transformed uranium concentration variable (`log_U_intake`).

```{r histograms, warning = FALSE, echo = FALSE}
histo <- ggplot(nn_data, aes(x = U_intake))+
  geom_histogram(binwidth = 50) +
  theme_minimal(base_size = 15)
histo
# we see the data is right-skewed with a large unimodal peak at the lowest uranium concentrations, so we perform a log-transform of the U_intake variable:
nn_data_logged <- nn_data %>% 
  mutate(log_U_intake = log(U_intake))
# log histogram
histo_log <- ggplot(nn_data_logged, aes(x = log_U_intake))+
  geom_histogram(binwidth = 0.5) +
  theme_minimal(base_size = 15)
histo_log
```

Then, wanting to re-visit and determine if there was any relationship between the log-transformed uranium intake and glucose, we re-created the scatter plot using those two variables, and created a linear regression model with a best fitting line for the scatter plot. The Pearson's correlation was `0.42`. The linear model produced a intercept of `60.82` and a slope of `15.12`.

```{r correlations, warning=FALSE, echo=FALSE}
# Pearson's correlation
correlation <- nn_data_logged %>%
  summarize(cor_gluc_U = cor(log_U_intake, Glucose))
correlation
# linear model for log of U_intake and Glucose
reg_model <- lm(Glucose ~ log_U_intake, data = nn_data_logged)
reg_model
```

```{r summary-plots, warning=FALSE, echo=FALSE}
# scatter plot with the regression line
sc_plot_log <- ggplot(nn_data_logged, aes(x = log_U_intake, y = Glucose))+
  geom_point(aes(col=Exp_Level)) +
  geom_abline(slope = 15.12, intercept = 60.82) +
  theme_minimal(base_size = 15)
sc_plot_log
```

Finally, to further summarize the relationship, we created a box and whisker plot of glucose grouped by exposure level (`Exp_Level`) to visualize any association.

```{r box-whisker-plot, warning=FALSE, echo=FALSE}
# box-whisker plot
boxWhisker <- ggplot(nn_data, aes(y = Glucose, color=Exp_Level)) +
  geom_boxplot() +
  labs(title = "Summary",
       caption = "Data from Zamora",
       y = "Glucose in mg/d"
       ) +
  theme_minimal(base_size = 15)
boxWhisker
```
\newpage

## Part 2 Questions

1. [1 mark] Include your work for Part I.

Please see all prior pages.

\newpage

2. [2 marks] Describe a quantity you will estimate as an outcome in your problem using probability
notation. Are you planning to calculate marginal probabilities? Conditional probabilities?

Conditional probabilities of glucose excretion level is high or low given high or low exposure.
Conditional probabilities of creatinine excretion level is high or low given high or low exposure. These measures were per day accumulations, since participants had samples taken once in a 24 hour period. 

P(High Glucose | High exposure)

P(High Glucose | Low exposure)

P(Low Glucose | High exposure)

P(Low Glucose | Low exposure)

P(High Creatinine | High exposure)

P(High Creatinine | Low exposure)

P(Low Creatinine | High exposure)

P(Low Creatinine | Low exposure)

\newpage

3. [3 marks] Describe the type of theoretical distribution that is relevant for your data.
i) What type of variable(s) are you investigating (continuous, categorical, ordinal, etc)?

We are investigating continuous variables.

ii) What theoretical distribution that we have talked about would potentially be appropriate to use with
these data (Normal, Binomial, Poisson. . . )

For the scope learned in this class, we would anticipate it is a normal distribution.  However, the paper explicitly stated that these variables did not model a normal distribution due to failing a Shapiro-Wilk statistical test of normality.

iii) Why is this an appropriate model for the data you are studying?

Continuous variables cannot be in a binomial or Poisson as it is not a success/failure outcome.

\newpage

4. [4 marks] Use the data you have to demonstrate a statistical concept from Part II of the course. Describe
the concept that you are demonstrating and interpret the findings. This may include code in R, a visual
of some kind and text interpretation.

We know from our summary statistics that the Glucose and Creatinine continuous response variables have fairly similar respective means and medians (Glucose mean of 96.08, median of 73.75; Creatinine mean of 9.712, median of 8.95), so we might suspect those variables to be normally distributed, since having equal mean and median is a property of that distribution. To help assess normality of these variables in our sample (n = 50), we generated histograms and QQ plots to assess the shape of the distributions. Next, we created a randomly generated normal distribution for each of the two variables using their respective means and standard deviations. Initially, we created these random distributions using a sample size equal to what was performed in our reference study. Observing that the distributions generated with such a small sample size were not necessarily normal distributions, we then created these same normal distributions, but used a larger sample size (n) of 200 to illustrate the Central Limit Theorem.


```{r normal-plots, echo = FALSE, warning = FALSE}
histo_glucose <- ggplot(nn_data, aes(x = Glucose))+
  geom_histogram(binwidth = 50) +
  theme_minimal(base_size = 15) +
  labs(
    title = "Observed Glucose Data",
    caption = "n=50"
  )
  
histo_glucose
histo_creat <- ggplot(nn_data, aes(x = Creatinine))+
  geom_histogram(binwidth = 1) +
  theme_minimal(base_size = 15) +
  labs(
    title = "Observed Creatinine Data",
    caption = "n=50"
  )
histo_creat
qq_plot_glucose <- ggplot(nn_data, aes(sample = Glucose)) + 
  stat_qq() +
  stat_qq_line() +
  theme_minimal(base_size = 15)+
  labs(
    title = "QQ Plot for Glucose normality"
  )
qq_plot_glucose
qq_plot_creat <- ggplot(nn_data, aes(sample = Creatinine)) + 
  stat_qq() +
  stat_qq_line() +
  theme_minimal(base_size = 15) + 
  labs(
    title = "QQ Plot for Creatinine normality"
  )
qq_plot_creat
```

```{r rnorm}
# G_mean = 96.08, sd = 80.4045, C_mean = 9.712, sd = 3.415
# glucose normals
glucose_normal_nSmall <- rnorm(50, 96.08, 80.4045)
glucose_normal_nBig <- rnorm(200, 96.08, 80.4045)
glucose_normal_nSmall <- data.frame(glucose_normal_nSmall)
glucose_normal_nBig <- data.frame(glucose_normal_nBig)
# creatinine normals
creatinine_normal_nSmall <- rnorm(50, 9.712, 3.415)
creatinine_normal_nBig <- rnorm(200, 9.712, 3.415)
creatinine_normal_nSmall <- data.frame(creatinine_normal_nSmall)
creatinine_normal_nBig <- data.frame(creatinine_normal_nBig)
# histograms
histo_gluc_norm_small <- ggplot(glucose_normal_nSmall, aes(glucose_normal_nSmall))+
  geom_histogram(binwidth = 10) +
  theme_minimal(base_size = 15) +
  labs(
    title = "Glucose generated normal Histogram",
    caption = "Small sample sizes (n=50)"
  )
histo_gluc_norm_small
histo_gluc_norm_big <- ggplot(glucose_normal_nBig, aes(glucose_normal_nBig))+
  geom_histogram(binwidth = 10) +
  theme_minimal(base_size = 15) +
  labs(
    title = "Glucose generated normal Histogram",
    caption = "Large sample sizes (n=200)"
  )
histo_gluc_norm_big
histo_creat_norm_nSmall <- ggplot(creatinine_normal_nSmall, aes(creatinine_normal_nSmall))+
  geom_histogram(binwidth = 1) +
  theme_minimal(base_size = 15) +
  labs(
    title = "Creatinine generated normal Histogram",
    caption = "Small sample sizes (n=50)"
  )
histo_creat_norm_nSmall
histo_creat_norm_nBig <- ggplot(creatinine_normal_nBig, aes(creatinine_normal_nBig))+
  geom_histogram(binwidth = 1) +
  theme_minimal(base_size = 15) +
  labs(
    title = "Creatinine generated normal Histogram",
    caption = "Large sample sizes (n=200)"
  )
histo_creat_norm_nBig
```

\newpage

## Part 3 Questions

### Question 1: Identify a statistical test to apply to your data (must be a concept we covered in part III of the course). In plain language, write the question you are trying to answer.

We'll do a two-sample t-test of mean Glucose levels (response variable), based on high and low uranium exposure (exposure).

\newpage

### Question 2: What assumptions are required by the method you chose in 5.1? Show how you assessed whether these assumptions are met by your dataset.

We feel comfortable running this test statistic since the collected data are, based on previous data project sections:

1. approximately normal,
2. sufficient sample size (n),
3. independent variables,
4. simple random samples,
5. and, mostly free of outliers.

When running our statistic, we'll exclude the two values that we suspect are possible outliers (Glucose >350)

\newpage

### Question 3: Explain why this test is appropriate for the data you have and the question you are trying to answer. Use at least one visualization technique and include both the output and the R code that generated it.

We're looking at two continuous variables, with Glucose being the first, and Uranium, but we've sorted Uranium categorically based on high or low exposure. As well, they're independent variables, and we want a parametric test statistic because the we don't have population level data for comparison, thus we use a two-sample t-test.

```{r t-test vis}

# uranium intake log transformed, Glucose left alone, removing 2 outliers of Glucose > 350
nn_data_logged_filtered <- nn_data_logged %>%
  filter( Glucose < 350)

# High and low uranium exposure for histogram data vis
nn_data_high <- nn_data_logged_filtered %>%
  filter(Exp_Level == "H")
nn_data_low <- nn_data_logged_filtered %>%
  filter(Exp_Level == "L")

# red = high, blue = low
comparison_histogram <- ggplot()+
  geom_histogram(data = nn_data_high, aes(x = Glucose, color = Exp_Level), alpha = 0.5, binwidth = 25) +
  geom_histogram(data = nn_data_low, aes(x = Glucose, color = Exp_Level), alpha = 0.5, binwidth = 25) +
  theme_minimal(base_size = 15) +
  labs(
    title = "Observed Glucose Data",
    caption = "Removed 2 outliers for the Uranium High exposure group where Glucose > 350"
  )
comparison_histogram
```

\newpage

### Question 4: Clearly state the null and alternative hypotheses for this test.

*Null hypothesis:* For the two groups (high and low uranium exposure), the mean Glucose level will be equal.
*Alternate hypothesis:* For the two groups (high and low uranium exposure), the mean Glucose level will not be equal.

\newpage

### Question 5: Include the R code you used to generate your results. Annotate your code to help us follow your reasoning.

```{r two sample t-test}

# Glucose is the response variable, compared by the categorical variable of Exp_Level, using the nn_data set but filtered first to remove two possible outliers.
t.test(Glucose ~ Exp_Level, data = nn_data_logged_filtered)
```

\newpage

### Question 6: Present your results in a clear summary. This should include both a text summary and a table or figure with appropriate labeling.

From our analysis, the two-sample t-test statistic was 3.2503, with a p-value of 0.0023, a 95% confidence interval for the difference of the means between 14.72 and 63.12 mg/Day.

#### Table 1:

|       | t-test statistic | p-value  | 95% CI (difference between means) | High Exposure,<br> Mean Glucose | Low Exposure,<br> Mean Glucose |
|-------|------------------|----------|-----------------------------------|--------------------------------|-------------------------------|
| Value | 3.2503           | 0.002338 | 14.72, 63.12                      | 99.32857                       | 60.41000                      |

\newpage

### Question 7: Interpret your findings. Include a statement about the evidence, your conclusions, and the generalizability of your findings.

Based on these results, we would say that the p-value of 0.0023 and 95% CI mean we have evidence to reject the null hypothesis, and that there is a significant difference in mean Glucose level between low- and high- uranium exposure. High glucose levels in urine is not typical physiologically in healthy individuals. This supports the idea that human health can be affected by uranium exposure, likely indicating kidney damage and injury associated with these elevated glucose levels. 

The dataset we used likely may not be generalized beyond the study population, since the study population consisted of individuals living in Canada, it was such a small sample size (n=50), and half of those individuals were those living in a specific city, while the other half were a rural community (a Nova Scotian village) living nearby to known uranium contaminated water sources. However, it still does provide useful evidence that there is an association of uranium exposure and kidney damage, that can inform future research for other target populations and communities dealing with contaminated water sources such as the Navajo Nation.

\newpage

#### Attributions

*Contributors:* All team members led the production of this manuscript and Matthew Reyes had primary responsibility for the composition. Vicki Ip contributed revisions and had primary responsibility for coordinating with our Graduate Student Instructor throughout the course. The corresponding author attests that all listed authors meet authorship criteria and that no others meeting the criteria have been omitted.
